Simulating High-Dimensional Multivariate Data

### 2021-02-25 010-background.Rmd
small changes up to 020

The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful timescale.

@@

\begin{equation}
(\#eq:pearson)
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}}.
\end{equation}

@@

*Correspondence among Pearson, Spearman, $\tau$ correlations*.
There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
For *bivariate normal vectors*, however, the correspondence is well-known:

\begin{equation}
(\#eq:convertKendall)
\rho_{P} = sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

@@

\begin{equation}
(\#eq:convertSpearman)
\rho_P = 2 \times sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

@@

\begin{equation}
F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}.
(\#eq:inverseCDF)
\end{equation}

@@

For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying

@@

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ provides $Y_i \sim F_i, \, \forall \, i$.

>>>>

The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in a practical timescale.

@@

\begin{equation}
(\#eq:pearson)
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ \mathrm{Var}(X)\mathrm{Var}(Y)\right]^{1/2}}.
\end{equation}

@@

*Correspondence among Pearson, Spearman, Kendall correlations*.
There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation. For *bivariate normal vectors*, however, the correspondence is well-known:

\begin{equation}
(\#eq:convertKendall)
\rho_{P} = \sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

@@

\begin{equation}
(\#eq:convertSpearman)
\rho_P = 2 \times \sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

@@

\begin{equation}
F_{i}^{-1} = \inf\{y:F_{i}(y) \geq u \}.
(\#eq:inverseCDF)
\end{equation}

@@

For any random vector ${\bf X}=(X_1, \ldots, X_d)^\top$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying

@@

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)^\top$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ provides $Y_i \sim F_i, \, \forall \, i$.

### 2021-02-25 020-simulation-algorithm.Rmd
small changes up to 020

1. Simulate a random vector $\bf Z$ with $d$ *independent* and *identical* standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,2,...,d$.

@@

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic. In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$. Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,2,\ldots,d, \; i \neq j$. For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies. For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].

@@

There is hope, however, as @GH02 also showed that replicating an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides evidence that our nearest positive semi-definite (PSD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

@@

1. Mapping step  
   i. Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.   
   or  
i$^\prime$. Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
2. Checking admissibility and, if needed, nearest correlation computation
    (i) Check that $R_X$ is a correlation matrix, a PSD matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.  
	Otherwise,  
(ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $R_X$, in the Frobenius norm.
3. NORTA step
	(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
	(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
	(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

@@

Yet this scheme still suffers similar issues as the original Pearson-matching NORTA algorithm regarding the increasing probability of encountering a non-admissible correlation matrix as dimension increases. In our experience, the *mapping step* for large $d$ almost always produces a $R_X$ that is not in $\Omega_P$ (or even a correlation matrix). The second step handles this issue by computing the nearest correlation matrix to $R_X$. In Section \@ref(package) we provide a basic description of how to quantify and potentially control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

@@

$r = 2 \sin \left(\frac{\pi}{6} \rho\right) \\
r = \sin \left(\frac{\pi}{2} \tau\right)$

@@

* Determine the input correlation, $\tilde{P}$, such that the resulting random samples using the NORTA algorithm have an estimated Pearson correlation that matches $P$

>>>>

1. Simulate a random vector $\bf Z$ with $d$ *independent and identically distributed* (iid) standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,...,d$.

@@

Matching specified Pearson correlation coefficients exactly (step 2 above), however, is problematic. In general, there is no closed-form correspondence between the components of the input $\Sigma_{\bf Z}$ and target $\Sigma_{\bf Y}$. Matching the correlations involves evaluating or approximating $\binom{d}{2}$ integrals of the form $EY_iY_j = \int \int y_i y_j f_{X}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$, for $i,j=1,\ldots, d, \; i \neq j$. For high-dimensional data, these evaluations are often too costly to enable feasible simulation studies. For low-dimensional problems, however, methods and tools exist to match Pearson correlations precisely [@Xia17] and the publicly available `nortaRA` R package [@Chen2001].

@@

There is hope, however, as @GH02 also showed that replicating an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides evidence that our nearest positive definite (PD) augmented approach will maintain reasonable accuracy if our input matching scheme returns an indefinite matrix, at least for the rank-based matching scheme described above.

@@

1. Mapping step
  (i) Convert the target Spearman correlation matrix $R_S$ to the corresponding MVN Pearson correlation $R_X$.
  or
  (i$^\prime$) Convert the target Kendall $\tau$ matrix $R_K$ to the corresponding MVN Pearson correlation $R_X$.
2. Check admissibility and, if needed, compute the nearest correlation matrix.
  (i) Check that $R_X$ is a correlation matrix, a positive [semi]definite matrix with 1's along the diagonal.   
	(ii) If $R_X$ is a correlation matrix, the input matrix is *admissible* in this scheme.
	Otherwise,
  (ii$^\prime$) Replace $R_X$ with the nearest correlation matrix $R_X$, in the Frobenius norm.
3. NORTA step
  (i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_X)$.  
  (ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$.  
  (iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$.

@@

Yet this scheme still suffers similar issues as the original Pearson-matching NORTA algorithm regarding the increasing probability of encountering a non-admissible correlation matrix as dimension increases. In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not in $\Omega_P$ (or even a correlation matrix). The second step handles this issue by computing the nearest correlation matrix to $R_X$. In Section \@ref(package) we provide a basic description of how to quantify and potentially control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

@@

$$
r = 2 \sin \left(\frac{\pi}{6} \rho\right) \\
r = \sin \left(\frac{\pi}{2} \tau\right)
$$

@@

* Determine the input correlation, $\tilde{R}$, such that the resulting random samples using the NORTA algorithm have an estimated Pearson correlation that matches $R$

### 2021-02-25 index.Rmd
small changes up to 020

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the `Bigsimr` Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package. This high-package provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Kendall, or Spearman correlation matrices. It also contains additional high-performance features, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
---

@@

```{r indexSetup, include=FALSE, cache=FALSE}
library(tidyverse)
library(knitr)
library(printr)
library(rmarkdown)
library(bookdown)

@@

```{r countWords, include=FALSE, eval=FALSE}
rmdFiles <- dir(pattern = 'Rmd$', ignore.case= T)
# rmdFiles <- rmdFiles[-grep('^_', rmdFiles)]
rmdFiles <- rmdFiles[-grep('references', rmdFiles)]
# rmdFiles <- rmdFiles[-grep('README', rmdFiles)]
# rmdFiles <- rmdFiles[-grep('article', rmdFiles)] ## after generating a single document
# wordcountaddin::word_count( rmdFiles[11] )
(counts <- sapply( rmdFiles, wordcountaddin::word_count, simplify = T ))
sum(counts)
```

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the `Bigsimr` Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package. This high-level package provides a flexible and scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Spearman, or Kendall correlation matrices. It also contains additional high-performance features, including multi-core algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
---

@@

```{r indexSetup, include=FALSE, cache=FALSE}
library(tidyverse)

@@

```{r countWords, include=FALSE, eval=FALSE}
rmdFiles <- dir(pattern = 'Rmd$', ignore.case = TRUE)
# rmdFiles <- rmdFiles[-grep('^_', rmdFiles)]
rmdFiles <- rmdFiles[-grep('references', rmdFiles)]
# rmdFiles <- rmdFiles[-grep('README', rmdFiles)]
# rmdFiles <- rmdFiles[-grep('article', rmdFiles)] ## after generating a single document
# wordcountaddin::word_count( rmdFiles[11] )
(counts <- sapply( rmdFiles, wordcountaddin::word_count, simplify = T ))
sum(counts)
```
